A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-10-24 doc/types-pldi2013/langdef.tex

The final element of the environtment is the {\em clobber set} $C$.  Our goal is to show that parallel Legion
programs can
be scheduled in a hierarchical (and therefore scalable) fashion, we require a formulation of operational semantics
that is both parallel and hierarchical.  Because Legion allows constrained nondeterminism (in the order in which
reductions are applied and in cases of relaxed coherence), it is possible for the heap accesses performed by
one expression's evaluation to observe interference from other concurrent evaluations.  The traditional way of
handling this in operational semantics \cite{XYZ} is to use a top-down approach with small-step semantics and
a global view of all
partially-evaluated expressions.  Rules are specified for which expression(s) are valid choices for the next small
step of execution.  This approach has the benefit of being constructive but the implicit global ``scheduler'' is
difficult to map into a hierarchical one in which decisions must necessarily be made with limited information.

@@

The complete operational semantics rules can be found in \cite{LegionTypes12} - we show the most interesting ones
in Figure~\ref{fig:semantics}.  The rules for the {\em read}, {\em write}, and {\em reduce} show how dynamic memory
traces are constructed and how the result of a read expression is relaxed when the location falls within the
clobber set.  Additionally, the write and reduce operations, with their two subexpressions, demonstrate how the
traces of subexpressions are interleaved within the environment (and clobber set) of the parent expression.

@@

Theorem~\ref{thm:effects} guarantees that a function call produces a value of the
correct type and does so within the regions of memory for which privileges are held.
It does not, however, address {\em coherence}, what concurrent operations on the regions are permitted.
An innovation in Legion is that programmers can control coherence on a per-region basis.  Intuitively,
the different coherence modes permit different interleavings of memory traces.  For example,
in addition to being an interleaving of the constituent traces, exclusive access requires interleaved memory traces to be:
\begin{itemize}
\item {\em coherent} - a read of location $l$ should see the effect of all writes and reductions to $l$
occurring earlier in the interleaved trace, and
\item {\em sequentially equivalent} - the value read by each read 
operation and the result of applying the interleaved trace to the original store should be the same as if each constituent trace
was applied to the store in a valid sequential execution of the program.
\end{itemize}
These requirements are relaxed for either a non-empty clobber set (representing
operations that may occur concurrently as part of evaluating other functions) or for coherence
modes other than {\em excl}.

@@

Figure~\ref{fig:validinterleave} formally defines {\em
valid\_interleave}, which we briefly explain.  Consider
traces $\oton{E}$.  The {\em exclusive location set} $L_{excl}$ is the
set of locations used in an operation with coherence mode $excl$ in
some trace $E_i$, excepting any locations in the clobber set $C$.
Similarly, the {\em atomic location set} $L_{atomic}$ is the set of
locations in an operation with coherence mode $atomic$, but no
operations with coherence mode $excl$, again excepting any location in
the clobber set.  The $coherent(S,L_1,L_2,E)$ test checks that all
reads of locations in $L_2$, and any read of a location $l \in L_1$
that comes after a write to $l$, return the value most recently written
to the store.  (If $L_2 = \emptyset$ initially, atomic reads without a preceding write are not 
checked, but are deferred to a point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
test verifies every read in $E'$ with a location in $L_1$ 
is coherent if operations are performed in 
program order and that the store that results from applying $E'$
matches the store resulting from applying the
constituent traces in program order for all locations in $L_1$.  Program order
is implicitly defined in the operational semantics, and is equivalent to a 
postorder traversal of an expression's tree.  For
the atomic location set, we allow the constituent traces to be
permuted, but all locations must be checked with the same permutation.

Flexibility in interleaving operations is the key to achieving parallel execution, but
implementing the precise criteria for checking a valid interleaving is
much too expensive.  We now consider practical mechanisms that generate valid interleavings and 
under what conditions these approaches may be used.

The first option is to serialize the evaluation of the subexpressions that
make up an expression and attempt to generate a valid interleaving by simply
concatenating the constituent traces in program order.  However, there is a
complication related to the non-determinism that is allowed for reads.  Although
each subexpression's trace will have been checked for coherence, the checking
of some operations may have been deferred due to non-empty
clobber sets.  It is possible that when an interleaving of these traces
is attempted in the parent expression's context, there may no valid interleaving
if the wrong guesses were made in the derivation of a subexpression.
The solution in the case of serial execution is to not expand the clobber
set for subexpressions, instead using the parent expression's clobber set.

Only a single rule is modified for this alternate execution - $e_{n+1}$ in
{\sc E-Call} uses the original $C$ instead of $C'$.  This change
can result in there being no terminating execution for an expression in
which the only terminating executions required a particular (non-program-)
ordering of subexpression evaluation.  However, the Legion core language never
guarantees parallel execution, so the evaluation of such an 
expression is unlikely to terminate even under the original clobber set.
We can, however, show that any terminating execution that results from using
the restricted clobber set is a valid execution under the original:

@@

The proof of this (and other lemmas in this section) is included in \cite{LegionTypes12}.
We now show that the concatenation of the memory traces from these serially
executed subexpressions is always a valid interleaving:

@@

\begin{proof}
First, we must show that $any\_interleave(E', \oton{E})$.  This is done by
induction on the length of $E_1$, followed by $E_2$, and so on.  Next, we must
show that $E'$ is coherent for both $L_{excl}$ and $L_{atomic}$.  This is done by
matching the structure of $E'$ with $\oton{E}$ and the use of the following
lemma: $apply(S, E_1 \dplus E_2) = apply(apply(S, E_1), E_2)$.  With the
question of coherence addressed, the sequential equivalence of $E'$ is trivial
for both the exclusive and atomic cases.
\end{proof}

>>>>

The final element of the environment is the {\em clobber set} $C$.  Our goal is to show that parallel Legion
programs can
be scheduled in a hierarchical (and therefore scalable) fashion, we require a formulation of operational semantics
that is both parallel and hierarchical.  Because Legion allows constrained nondeterminism (in the order in which
reductions are applied and in cases of relaxed coherence), it is possible for the heap accesses performed by
one expression's evaluation to observe interference from other concurrent evaluations.  The traditional way of
handling this in operational semantics \cite{XYZ} is to use a top-down approach with small-step semantics and
a global view of all
partially-evaluated expressions.  Rules are specified for which expression(s) are valid choices for the next small
step of execution.  This approach has the benefit of being constructive but the implicit global ``scheduler'' is
difficult to map into a hierarchical one in which decisions must necessarily be made with limited information.

@@

The complete operational semantics rules can be found in \cite{LegionTypes12} - we show the most interesting ones
in Figure~\ref{fig:semantics}.  The rules for the {\em read}, {\em write}, and {\em reduce} show how dynamic memory
traces are constructed and how the result of a read expression is relaxed when the location falls within the
clobber set.  Additionally, the write and reduce operations, with their two subexpressions, demonstrate how the
traces of subexpressions are interleaved within the environment (and clobber set) of the parent expression.
(A formal definition of the {\em valid\_interleave} predicate can be found in Figure~\ref{fig:validinterleave}, and
will be discussed in Section~\ref{sec:coherence}.)

@@

Theorem~\ref{thm:effects} guarantees that a function call produces a value of the
correct type and does so within the regions of memory for which privileges are held.  In fact, this results holds
even if the {\em valid\_interleave} constraint is weakened to just require an arbitrary interleaving of the memory
traces of subexpressions (i.e. the {\em any\_interleave} predicated defined in Figure~\ref{fig:opsecfns}).  The 
additional constraints on {\em valid\_interleave} in Figure~\ref{fig:validinterleave} address the coherence of 
heap accesses and allow for the per-region relaxation of coherence that is introduced in Legion.

To determine whether an interleaving of two or more memory traces is valid, we consider three sets of addresses:
\begin{itemize}
\item exclusive locations ($l \in L_{excl}$) are those which have at least once access in exclusive mode in the traces
and are not in the clobber set.  For these locations, we require the standard {\em sequential consistency} - all
reads to these locations see the effect of previous writes and reductions, and the resulting state of the store is
as if all writes and reductions were applied from each trace in order
\item atomic locations ($l \in L_{atomic}$) are those which have at least one access in atomic mode in the traces and
are in neither $L_{excl}$ nor the clobber set.  For these locations, we allow permutations of the original
subexpression trace order.
\item for locations that have only access in {\em simult} mode or are in the clobber set, no constraints are
enforced.  The validity of the interleaving of these accesses will be performed in the an enclosing expression.
\end{itemize}

@@

Although it is our goal to enable parallel execution whenever possible, it will help to briefly
consider serialized execution of a Legion program.  We will show that a serial execution trivially
satisfies the interleaving criteria required by the Legion operational semantics.  In addition to
serial execution being the fallback behavior of the hierarchical schedule, our proof of the
soundness of the parallel scheduling will depend on this result.

Serial execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
instead, and by forming ``interleaved'' traces by simply concatenating the subexpressions' traces.  By ignoring
the coherence modes, the clobber set remains empty, making the result of all {\em read} expressions fully
determined.  The following lemma and theorem (proofs of which can be found in \cite{LegionTypes12}) show that the
value and memory trace that result from a serialized execution of an expression are always valid executions of
Legion programs.

@@

@@

CURRENT EDIT POINT
